Ant Transition Matrix

### 2020-10-21 Ant Matrix_TOM.Rmd
progress on new file

```{r}
#import the data
cactus <- read.csv("/Users/alicampbell/Desktop/cholla.csv", header = TRUE)
#str(cactus)
```

##Data Filtering
```{r}
## function for the volume of a cone

@@

```{r}
##Year T
cactus <- data.frame(lapply(cactus,      # Convert data with toupper function
                              function(variables) {
                                if (is.character(variables)) {
                                  return(toupper(variables))
                                } else {
                                  return(variables)
                                }
                              }),
                       stringsAsFactors = FALSE)
cactus_new <- as.data.frame(gsub("LIOM", "L",cactus$Antsp_t))
cactus_new <- as.data.frame(gsub("CREM", "C",cactus$Antsp_t))
cactus_new <- as.data.frame(gsub("FOR", "F",cactus$Antsp_t))

##Year T1
cactus <- data.frame(lapply(cactus,      # Convert data with toupper function
                              function(variables) {
                                if (is.character(variables)) {
                                  return(toupper(variables))
                                } else {
                                  return(variables)
                                }
                              }),
                       stringsAsFactors = FALSE)
cactus_new <- as.data.frame(gsub("LIOM", "L",cactus$Antsp_t1))
cactus_new <- as.data.frame(gsub("CREM", "C",cactus$Antsp_t1))
cactus_new <- as.data.frame(gsub("FOR", "F",cactus$Antsp_t1))

#unique(cactus$Antsp_t)
```

```{r}
#Bin all ant species in year t to a value 1 - 5 without taking away the species denotation
cactus$bin1 <- 0 #blank column 
for(i in seq_len(nrow(cactus))){
  if(cactus$Antsp_t[i] == "LIOM"){cactus$bin1[i] <- 1}
  else if(cactus$Antsp_t[i] == "L"){cactus$bin1[i] <- 1}
  else if(cactus$Antsp_t[i] == "LIOM "){cactus$bin1[i] <- 1}
  else if(cactus$Antsp_t[i] == "CREM"){cactus$bin1[i] <- 2}
  else if(cactus$Antsp_t[i] == "C"){cactus$bin1[i] <- 2}
  else if(cactus$Antsp_t[i] == "FOR"){cactus$bin1[i] <- 3}
  else if(cactus$Antsp_t[i] == "F"){cactus$bin1[i] <- 3}
  else if(cactus$Antsp_t[i] == "VACANT"){cactus$bin1[i] <- 4}
  else if(cactus$Antsp_t[i] == ""){cactus$bin1[i] <- 4}
  else{cactus$bin1[i] <- 5}
}
#Bin all ant species in yeay t1 to a value 1 - 5 without taking away species denotation
cactus$bin2 <- 0 #blank column 
for(i in seq_len(nrow(cactus))){
  if(cactus$Antsp_t1[i] == "LIOM"){cactus$bin2[i] <- 1}
  else if(cactus$Antsp_t1[i] == "L"){cactus$bin2[i] <- 1}
  else if(cactus$Antsp_t1[i] == "LIOM "){cactus$bin2[i] <- 1}
  else if(cactus$Antsp_t1[i] == "CREM"){cactus$bin2[i] <- 2}
  else if(cactus$Antsp_t1[i] == "C"){cactus$bin2[i] <- 2}
  else if(cactus$Antsp_t1[i] == "FOR"){cactus$bin2[i] <- 3}
  else if(cactus$Antsp_t1[i] == "F"){cactus$bin2[i] <- 3}
  else if(cactus$Antsp_t1[i] == "VACANT"){cactus$bin2[i] <- 4}
  else if(cactus$Antsp_t1[i] == ""){cactus$bin2[i] <- 4}
  else{cactus$bin2[i] <- 5}
}
```

>>>>

```{r read_data}
#import the data
cactus <- read.csv("cholla_demography_20042019.csv", header = TRUE)
#str(cactus) ##<- problem: antcount is a factor
#levels(cactus$Antcount_t);levels(cactus$Antcount_t1)
## drop the offending rows -- this is a 2019 data entry problem
cactus <- cactus[-c(which(cactus$Antcount_t=="2-Jan"),which(cactus$Antcount_t1=="2-Jan")),]
cactus$antcount_t <- as.numeric(as.character(cactus$Antcount_t))
cactus$antcount_t1 <- as.numeric(as.character(cactus$Antcount_t1))
```

##Data Filtering
```{r volume}
## function for the volume of a cone

@@

```{r year_t_ants}
## assign ant counts of zero as vacant
cactus$Antsp_t[cactus$antcount_t==0] <- "vacant"
# here are the ordered levels of the current variable
Antsp_t_levels <- levels(cactus$Antsp_t)
# here is how I would like to collapse these into fewer bins -- most will be "other"
ant_t_levels <- rep("other",times=length(Antsp_t_levels))
## crem levels - elements 5,8,9
ant_t_levels[c(5,8,9)] <- "crem"
## liom levels - elements 15,18,19,20,21
ant_t_levels[c(15,18,19,20,21)] <- "liom"
## forelius levels - 12,13
ant_t_levels[c(12,13)] <- "for"
## vacant - 31
ant_t_levels[31] <- "vacant"
## create new variable merging levels as above
cactus$ant_t <- factor(cactus$Antsp_t,levels=Antsp_t_levels,labels=ant_t_levels)
summary(cactus$ant_t) ## there are too many "others". check these out.
filter(cactus,ant_t=="other")
## first, there are no ant data from 2007:
cactus$Antcount_t[cactus$Year_t==2007]
#so give these observations NA for ant status:
cactus$ant_t[cactus$Year_t==2007] <- NA
# also, if a plant was new in year t+1 it's year t ant status should be NA
cactus$ant_t[cactus$Newplant==1] <- NA
# plots 7 and 8 were added in 2011, so their year_t==2010 ant status should be NA
cactus$ant_t[cactus$Year_t==2010 & cactus$Plot==7] <- NA
cactus$ant_t[cactus$Year_t==2010 & cactus$Plot==8] <- NA
# check to see how many others this leaves -- still a lot
summary(cactus$ant_t)
## another 2019 data entry problem: new plants were not given Newplant==1 and their Survival is 0 (should be NA)
## for now here is my workaround: year t ==2018, height_t ==NA, height_t1 !=NA
cactus$ant_t[cactus$Year_t==2018 & is.na(cactus$Height_t) & !is.na(cactus$Height_t1)] <- NA
## finally there are these plants with Antsp_t=="" for all kinds of reasons but bottom line is that we don't have ant status
cactus$ant_t[cactus$Antsp_t==""]<-NA
## inspect the "others" visually
cactus %>% filter(ant_t=="other") %>% select(Antsp_t,ant_t)
## I am satisfied that everything assigned other is correctly assigned

```{r ant_time_series}
## plot a time series of ant frequency by year
ant_freq_ts <- cactus %>% 
  select(Year_t,ant_t) %>% 
  drop_na() %>% 
  group_by(Year_t,ant_t) %>% 
  summarise(n = n()) %>% 
  mutate(freq = n / sum(n))

plot(ant_freq_ts$Year_t[ant_freq_ts$ant_t=="vacant"],ant_freq_ts$freq[ant_freq_ts$ant_t=="vacant"],type="b",ylim=c(0,1),pch=16,cex=1.5,
     xlab="Year",ylab="Frequency")
legend("topleft",legend=c("Vacant","Liom.","Crem.","For.","Other"),pch=16,col=c("black","blue","red","green","pink"))
points(ant_freq_ts$Year_t[ant_freq_ts$ant_t=="liom"],ant_freq_ts$freq[ant_freq_ts$ant_t=="liom"],type="b",pch=16,cex=1.5,col="blue")
points(ant_freq_ts$Year_t[ant_freq_ts$ant_t=="crem"],ant_freq_ts$freq[ant_freq_ts$ant_t=="crem"],type="b",pch=16,cex=1.5,col="red")
points(ant_freq_ts$Year_t[ant_freq_ts$ant_t=="for"],ant_freq_ts$freq[ant_freq_ts$ant_t=="for"],type="b",pch=16,cex=1.5,col="green")
points(ant_freq_ts$Year_t[ant_freq_ts$ant_t=="other"],ant_freq_ts$freq[ant_freq_ts$ant_t=="other"],type="b",pch=16,cex=1.5,col="pink")

```{r year_t1_ants}
## assign ant counts of zero as vacant
cactus$Antsp_t1[cactus$antcount_t1==0] <- "vacant"
## repeat for t1 -- these indices are different
Antsp_t1_levels <- levels(cactus$Antsp_t1)
# here is how I would like to collapse these into fewer bins -- most will be "other"
ant_t1_levels <- rep("other",times=length(Antsp_t1_levels))
## crem levels 
ant_t1_levels[c(8,9,10)] <- "crem"
## liom levels 
ant_t1_levels[c(19,20,21,22)] <- "liom"
## forelius levels 
ant_t1_levels[c(13,14,15)] <- "for"
## vacant 
ant_t1_levels[32] <- "vacant"
## create new variable merging levels as above
cactus$ant_t1 <- factor(cactus$Antsp_t1,levels=Antsp_t1_levels,labels=ant_t1_levels)
summary(cactus$ant_t1) ## there are too many "others" -- same deal as above
## first, there are no ant data from 2007 so give these observations NA for ant status:
cactus$ant_t1[cactus$Year_t1==2007] <- NA
# check to see how many others this leaves -- still a lot
summary(cactus$ant_t1)
## another 2019 data entry problem: new plants were not given Newplant==1 and their Survival is 0 (should be NA)
## for now here is my workaround: year t ==2018, height_t ==NA, height_t1 !=NA
cactus$ant_t[cactus$Year_t==2018 & is.na(cactus$Height_t) & !is.na(cactus$Height_t1)] <- NA
## finally there are these plants with Antsp_t=="" for all kinds of reasons but bottom line is that we don't have ant status
cactus$ant_t[cactus$Antsp_t==""]<-NA
## inspect the "others" visually
cactus %>% filter(ant_t=="other") %>% select(Antsp_t,ant_t)
## I am satisfied that everything assigned other is correctly assigned
